% =============================================================
% Section 5: Discussion
% =============================================================
\section{Discussion}\label{sec:discussion}

The forward PINN here matches the architecture, training schedule
and physics setup of \citet{Patel2024pinn} to enable a direct
field-accuracy comparison. Two changes
account for the improvement from $\mathrm{RL2}=28.06\%$ in their
Table~2 to $\mathrm{RL2}=0.58\%$ in our
Eq.~\eqref{eq:fwd-metrics}. The first is the doubled number of L-BFGS
iterations ($30{,}000$ versus $15{,}000$). The second is the greedy adaptive
resampling of Sec.~\ref{sec:fwd-pinn} (with resampling period
$1{,}000$ iterations versus the $100$ of \citet{Patel2024pinn},
so that each resampling event uses a residual field evaluated
after more iterations). The frequency-domain Schwarzschild
eigenvalue PINN of \citet{Cornell2022pinnqnm} reaches sub-permille
accuracy on the same $\ell = 2$ fundamental mode by training
$\omega$ as a complex parameter alongside the network weights, but
that result is in the eigenvalue class and not directly comparable
to our time-domain post-hoc fit. \citet{Luna2022teukolskypinn}
enforce normalisation boundary conditions hard via an output
transform on a time-domain Teukolsky PINN; we use soft enforcement
through the loss, and a hard output-transform variant of the
forward setup of Sec.~\ref{sec:fwd-pinn} is a natural follow-up.
The inverse PINN of Sec.~\ref{sec:inv-pinn}, which jointly
recovers the black-hole mass and a single-mode ringdown template
from noisy field observations, has no precedent in the
PINN-for-QNM literature. We isolate the
$A_{\mathrm{ring}}$--$\tau$ degeneracy of the template loss
(Sec.~\ref{sec:res-inv} and Table~\ref{tab:tau-degeneracy}),
adopt a reporting convention (Sec.~\ref{sec:reporting}) that
extracts $\tau$ from the converged network field rather than from
the learned scalar, and replace the fixed-window QNM fit by the
plateau-based estimators M4 and M5, whose
in-window scatter also serves as a quality flag. The
six-variant ablation of Table~\ref{tab:variants} isolates the
effects of the ringdown weight $\lambda_{\mathrm{ring}}$, the
ringdown lower bound $\tring$, and the addition of an overtone
in the template.
The curriculum-learning extension of Sec.~\ref{sec:cur-pinn}
addresses a different question, namely whether splitting the temporal
domain into smaller windows lowers the full-domain forward error.
The answer is no.
Table~\ref{tab:cur-summary} shows that, while the
$W_{1} = [0,25]\,M$ window of the 2-window curriculum reaches
RMSD $5.2\times 10^{-4}$ (the lowest sub-domain RMSD recorded
anywhere in this work), the $W_{2}$ stitching error and the
late-time error growth offset the local gain, leaving the
full-domain RMSD slightly worse than the single-window
greedy-adaptive baseline; the 3-window variant compounds the
stitching error monotonically across windows, and the
soft-overlap variant performs worse still. The curriculum
literature \citep{Krishnapriyan2021failure,Bengio2009curriculum}
predicts gains for problems whose loss landscape is dominated by
late-time stiffness once the network has been forced to learn
the full domain at once. In our setup, the residual-greedy
adaptive sampler of Sec.~\ref{sec:fwd-pinn} already concentrates
$0.3\,N_r$ of the collocation budget on the high-residual
late-time band within a single network, and so absorbs the same
benefit a curriculum would provide without paying the per-stitch
interpolation cost. We expect the curriculum strategy to be
worthwhile when the time domain is long enough that even a
fully-greedy single-window PINN cannot cover the late-time decay
range with adequate collocation density (e.g.\ for
$t_{\max} \gtrsim 5\,t_{\mathrm{tail}}$, where the QNM regime,
the late tail, and the Sommerfeld arrival time all enter the
same training window). On the present
$[0,50]\,M$ domain a single greedy PINN suffices.
The analysis of Sec.~\ref{sec:res-inv} (Table~\ref{tab:tau-degeneracy})
shows that the optimiser drifts along the amplitude--damping-time
degeneracy of Eq.~\eqref{eq:loss-ring} during L-BFGS, leaving the
learned
$\tau$ ten to twenty percent below the Leaver value while the
network field $\Phi_\theta(\xq,t)$ retains the physical decay
rate; the reporting convention of Sec.~\ref{sec:reporting}
therefore quotes the post-hoc M2 wide-window value of $\tau$ on
$\Phi_\theta$, which recovers $\tau$ to within a few percent on
the same runs. We expect the same convention to apply to any
inverse PINN whose template loss admits a comparable
amplitude--damping-time degeneracy.

The M4 plateau standard deviation $\sigma_\omega$ is, by
definition, the spread of $\omega$ across the plateau, which we
take as an estimate of the fit-window systematic uncertainty inside
the fit window. Across the six inverse variants it
discriminates cleanly between configurations on which the
two-mode NLS finds a stable minimum
($\sigma_\omega/\omega \le 1\%$ for variants A, D, E) and
configurations on which it does not
($\sigma_\omega/\omega \approx 0.44$ for baseline and $0.34$ for C,
undefined for B; see Table~\ref{tab:inv-summary}). A small
$\sigma_\omega$ certifies the estimate; a large or undefined
$\sigma_\omega$ rejects it.
The 2-D plateau M5 generalises the same idea to the joint
$(t_0,\tend)$ plane (Fig.~\ref{fig:fwd-plateau-m5}) and is needed
when the upper edge of the simulation window approaches
$t_{\mathrm{tail}}$, as in extensions to longer time domains.

The forward PINN reaches $\omega$ accuracy $0.029\%$ on a
noiseless setup; the best inverse variant reaches $0.30\%$ on the
$1\%$-noise setup. The factor of ten gap is consistent with the
information content of the two training setups. The forward problem
is constrained by an exact analytic IC plus a known PDE on the full
simulation domain, with no observational uncertainty; the inverse
problem retains the PDE constraint but adds a data-fit term and a
template-fit term, on $\Nobs = 500$ samples
restricted to $\xq = 10\,M$ and corrupted at the percent level.
A first-principles Cram\'er--Rao bound on $(M,\omega,\tau)$ given
$N_{\mathrm{obs}}$, $\sigma_{\mathrm{noise}}$ and the ringdown
SNR would settle the question of whether the residual
$0.30\%$ is dominated by training systematics or by intrinsic
information loss, and is a natural quantitative follow-up
of this work.

Translating variant D into observer units fixes the precision of
the inverse pipeline relative to current detector capabilities.
For a
Schwarzschild remnant of mass $M = 65\,M_\odot$, comparable to the
final-mass scale of GW150914 \citep{LIGO2016ringdown}, the
fundamental $\ell{=}m{=}2$ frequency
$f_{220} = M\omega_0 / (2\pi M_{\mathrm s})
\approx 186\,\mathrm{Hz}$ and the damping time
$\tau_0 \approx 3.60\,\mathrm{ms}$ follow from the Leaver values
$M\omega_0 = 0.3737$ and $\tau_0/M = 11.241$ used throughout this
paper, with $M_{\mathrm s} = M\,(GM_\odot/c^3)
\approx 3.20\times 10^{-4}\,\mathrm{s}$. The $0.30\%$
fractional uncertainty on $\omega$ achieved by variant D maps to
$\sigma_{f_{220}} \approx 0.56\,\mathrm{Hz}$, and the $1.6\%$
uncertainty on $\tau$ maps to
$\sigma_{\tau_0} \approx 58\,\mu\mathrm{s}$. Both are at least an
order of magnitude smaller than the statistical uncertainties on
$(f_{220},\tau_{220})$ reported for GW150914-class events at
ringdown signal-to-noise $\rho_{\mathrm{ring}} \sim 8$--$10$
\citep{LIGO2016ringdown,Isi2019tests}, so on a single event the
inverse-PINN precision floor is well below the data-statistical
floor; the implication is that, at current sensitivities, the
limiting uncertainty on a single-event spectroscopy test will
remain the data noise rather than the network systematics. This balance is
expected to invert at the per-mille statistical sensitivities
forecast for next-generation detectors \citep{Berti2007ringdown},
at which point the forward--inverse precision gap will become
a quantitative limiting factor.

Six limitations of this study fall into two groups, three
scope choices and three substantive. The scope choices are the fixed Schwarzschild background, the
single-channel quadrupole wavefunction $\Phi(\xq,t)$, and the
single noise realisation per variant. Each lifts to a known
extension. Kerr follows by substituting the Teukolsky operator for the
Zerilli operator, multi-channel by training one head per
harmonic, and the statistical scope by running the inverse
pipeline at $\ge 30$ noise seeds and reporting the seed-to-seed
standard deviation alongside the plateau $\sigma_\omega$. The
substantive limitations are the
$A_{\mathrm{ring}}$--$\tau$ degeneracy itself, which the
reporting convention works around but does not remove and which
is a property of any single-mode template; the marginal benefit
of the two-mode template (variant E), whose additional learnable
parameters slightly improve $\tau$ but penalise $M$ and $\omega$,
leaving the choice of $\lambda_{\mathrm{ring}}$ and the relative
weighting of overtone modes without a quantitative criterion; and the
ill-conditioning of ESPRIT (M3) on a $\Phi_\theta(\xq,t)$ whose
dominant component is already a clean damped sinusoid, which
prevents M3 from contributing to the primary numbers and which
we report in Appendix~\ref{app:esprit} only.

A related scope point concerns the overtone itself.
The M4 template of Eq.~\eqref{eq:two-mode} fits the fundamental
and the first overtone simultaneously, and we credit this
two-mode structure for the sub-percent fundamental-mode accuracy
quoted in Sec.~\ref{sec:res-fwd} (the Giesler-style mechanism: the
overtone slot absorbs $n=1$ contamination that would otherwise
bias a single-mode template; see \citet{Giesler2019overtones}).
Reporting $(M\omega_1,\tau_1/M)$ as a separate result, however,
is not warranted in the present setup: applied across all
forward, inverse and curriculum variants, the M4 plateau-mean
overtone recovers $M\omega_1$ no better than $\sim\!20\%$ in
the most favourable window choice, regardless of estimator
hyperparameters. The proximate cause is that our
Gaussian wave-packet initial data of Sec.~\ref{sec:fwd-problem}
excites $n=1$ at far lower amplitude than the
numerical-relativity merger remnants on which the Giesler
recipe was demonstrated. Reproducing Giesler-style overtone
resolution would therefore require initial data
chosen to excite higher overtones (or a binary-merger-conditioned
training distribution), which is a natural target for
follow-up work and is outside the scope of the present
methods-paper benchmark. The currently active controversy over
whether the GW150914 overtone constitutes a genuine detection
\citep{Cotesta2022overtones,Isi2021priors} provides additional
motivation for treating quoted overtone parameters with care.

A natural next step is to recast the inverse problem as a
Bayesian posterior over $(M,\omega,\tau)$ with the trained PINN
as the forward operator, following \citet{Isi2021priors}: this
would produce posterior credible
intervals on the primary parameters that combine the
network-systematic uncertainty (which the plateau diagnostic
estimates) with the data-statistical uncertainty (which the
deterministic point estimate does not). Combined with a
Cram\'er--Rao baseline, such a posterior would close the loop
between the deterministic plateau methodology developed here and
the probabilistic framework that gravitational-wave
ringdown analyses ultimately need. On the training-dynamics
side, the static loss-weight vector
$\boldsymbol{\lambda}=(100,100,100,1,100,1,1)$ inherited from
\citet{Patel2024pinn} could be replaced by the learning-rate
annealing of \citet{Wang2021gradientpath}, by the
NTK-eigenvalue-based weighting of \citet{Wang2022ntk}, by the
per-collocation-point soft-attention weights of
\citet{McClenny2023sapinn}, or by the causal
time-weighting of \citet{Wang2022causal}, the latter being
particularly natural for the time-domain ringdown setting in
which the network must resolve the early-time pulse before the
late-time decay can be trusted; and the soft Sommerfeld boundary
enforcement of Sec.~\ref{sec:fwd-pinn} could be replaced by the
hard output-transform construction of \citet{Luna2022teukolskypinn}.

The Fourier neural operator surrogate of
Sec.~\ref{sec:res-fno} differs from the three PINN modalities
in the unit of work that the network amortises. Each PINN
trains on one Zerilli problem and feeds its converged field
into the post-hoc QNM extractors. The FNO trains once on
$1{,}000$ initial-data draws and then predicts the full
spacetime field for any in-distribution draw with a single
forward pass, after which the same extractors of
Sec.~\ref{sec:qnm-methods} apply unchanged. On the canonical
BH the v4 FNO recovers $\omega$ to $0.11\%$ and $\tau$ to
$0.10\%$, a factor of $\sim\!6$ in $\omega$ and $\sim\!20$ in
$\tau$ larger than the FD reference's own extraction error
($0.019\%$ and $0.005\%$ respectively); the residual gap is
set by the bulk $9.30\times 10^{-4}$ field RMSD of the
surrogate, not by the M4 algorithm. On the $100$-BH
in-distribution population the FNO and the FD reference are
statistically indistinguishable at median $\omega$ error
$0.343\%$ versus $0.335\%$ and median $\tau$ error $0.940\%$
versus $0.661\%$, separated by less than the binomial standard
error on $n \approx 60$ valid draws. At the larger observer
position $\xq = 10\,M$ neither source returns a valid M4
plateau, locating the extraction failure in the joint
window--observer geometry rather than in the surrogate. The
five modalities therefore sit at distinct points on a
precision--amortisation plane. The forward PINN reaches
$0.04\%$ in $\omega$ at the cost of one bespoke L-BFGS run per
BH, the inverse PINN reaches $0.30\%$ while simultaneously
recovering $M$, the curriculum PINN improves the per-window
error on the earliest sub-domain by $\sim\!37\%$ without
improving the global extraction, the end-to-end FNO reaches
$0.34\%$ on the population median with a single forward pass
per BH, and the hybrid coarse-FD + neural-residual surrogate
matches the FD reference's own M4 plateau floor at the
$\xq = 2\,M$ observer at a smaller per-event inference cost
than the end-to-end FNO and with a network almost two orders of
magnitude smaller.
The end-to-end FNO matches the inverse-PINN observation-noise
floor of $1\%$ Gaussian noise at a per-BH cost two to three
orders of magnitude smaller than any per-BH PINN.

The hybrid coarse-FD + neural-residual surrogate of
Sec.~\ref{sec:hyb-method} occupies a third position distinct
from both the per-BH PINNs and the end-to-end FNO. The classical
coarse-FD solve on $(\Delta x,\Delta t)_{\mathrm{c}} = (0.8,0.4)\,M$
at refinement ratio $k = 4$ --- one sixty-fourth of the fine-FD
cost --- carries the wave propagation, and the 2-D FNO
$\mathcal{H}_{\theta}$ predicts only the discretisation residual
of this deliberately weak prior, trained from a label-free
Richardson extrapolation of two coarse solves rather than from
the fine-FD field and gated to the under-resolved wavefronts by
a parameter-free gradient gate. This lets a $1.13\times 10^{6}$
parameter network (almost two orders of magnitude smaller than the
end-to-end FNO of Sec.~\ref{sec:fno-method}) cut the canonical-BH
bulk-field RMSD from $6.95\times 10^{-3}$ on the coarse-up
baseline to $6.85\times 10^{-4}$, and the population MSE across
the 100-BH test set by a factor of $\sim\!79$. The gain is
structural rather than incidental: the coarse-prior error is
strongly correlated with the prior gradient (log--log Pearson
$r = 0.917$), so the gradient gate routes the network's capacity
onto the wavefronts where the prior fails --- a $12\times$ error
reduction in the top gradient decile --- while holding its
residual speckle in the smooth interior and causal exterior to
the $10^{-6}$ level, three orders of magnitude below the
correction. The QNM-extraction story at the observer
$\xq = 2\,M$ is the
opposite of the field story. All three sources --- hybrid,
coarse-up baseline, and fine FD reference --- hit the same M4
plateau floor on the canonical BH and on the median of the
100-BH population (median $|\Delta\omega/\omega| \approx 0.05\%$
on both hybrid and baseline, against $0.014\%$ on the fine FD
reference), so the hybrid
is not the median winner and shows no QNM benefit at this
observer; the off-floor outliers in the population are
ill-conditioned single-window fits that appear on the fine FD
reference as well, not a hybrid effect. The hybrid surrogate
therefore sits at the inference-fastest, smallest-network corner
of the precision--amortisation plane, and its field-RMSD gain
over the coarse-up baseline does not translate into a median QNM
gain at this observer because the coarse-up extraction is
already at the M4 plateau floor.

The v4 FNO ablation of Sec.~\ref{sec:res-fno} also yields a
methodological observation transferable to the operator-learning
literature on hyperbolic PDEs. The bulk-field RMSD across the
three versions of the surrogate (Table~\ref{tab:fno-versions})
is not monotone with the QNM extraction accuracy. The v2
configuration with the shortest training window has the lowest
bulk RMSD but the worst QNM fit, and the v4 configuration has
a slightly higher bulk RMSD than v3 but two orders of magnitude
better canonical-BH QNM accuracy. The two new v4 loss terms are
what convert v3 into v4. They are direct $L^{2}$ supervision on
the 1-D observer slice $\Phi(\xq,t)$ that the QNM extractor
reads, and a ramp-weighted field $L^{2}$ that puts $\sim 50\times$
more pressure on the ringdown tail than on the prompt pulse.
They make plain that the figure of merit for an
operator surrogate trained for ringdown spectroscopy is not the
gridwise RMSD but the downstream estimator accuracy on the slice
that the post-processing pipeline consumes. A natural extension
is to fold the inverse problem of Sec.~\ref{sec:res-inv} into
the operator setting itself: a network that learns the inverse
operator $\mathcal{G}^{-1}_{\theta}\colon \Phi(\xq,t) \mapsto M$
on the same training distribution would amortise the per-event
inverse PINN training cost in the same way that the v4 forward
FNO amortises the per-event forward PINN training cost, and the
plateau diagnostic of Sec.~\ref{sec:plateau} would carry over
to give an a-posteriori estimate of the population-level
systematic uncertainty on $M$.
